Predicting complex diseases: performance, robustness and ..utility..(impact?)

### 2018-02-12 3-simus.Rmd
htmls
save

```{r, message=FALSE}
library(tidyverse)
```

@@

    cmsa.logit <- big_CMSA(FUN = big_spLogReg, feval = AUC,
                           X = G, y.train = pheno.all[ind.train], 
                           ind.train = ind.train, 
                           covar.train = covar.all[ind.train, , drop = FALSE],
                           alpha = 0.5, dfmax = 20e3, 
                           ncores = nb_cores())

    preds <- predict(cmsa.logit, X = G, ind.row = ind.test, 
                     covar.row = covar.all[ind.test, , drop = FALSE])
  })[3]

  tibble(
    method   = method, 
    eval     = list(cbind(preds, pheno.all[ind.test])),
    timing   = timing,
    nb.preds = sum(cmsa.logit != 0)
  )
}
```

@@

n <- nrow(G)
n.train <- 6e3
ind.HLA <- snp_indLRLDR(CHR, POS, subset(LD.wiki34, ID == "hild12"))
```

@@

### Simulations n°1 (with T-Trees)

```{r}
params.grid1 <- expand.grid(
  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = 0.8, 
  par.model  = c("simple", "fancy"),
  num.simu   = 1:5,
  stringsAsFactors = FALSE
) 
```

```{r, eval=FALSE}
if (!dir.exists("results1")) dir.create("results1")

@@

  # Split in training/test sets
  ind.train <- sort(sample(n, size = n.train))
  ind.test <- setdiff(1:n, ind.train)

@@

### Simulations n°2 (without T-Trees, with logit-triple)

```{r}
params.grid2 <- expand.grid(
  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = c(0.5, 0.8), 
  par.model  = c("simple", "fancy"),
  num.simu   = 1:20,
  stringsAsFactors = FALSE
) 
```

```{r, eval=FALSE}
if (!dir.exists("results2")) dir.create("results2")

@@

### Simulations n°3 (without logit-triple)

```{r}
params.grid3 <- expand.grid(
  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = c(0.5, 0.8), 
  par.model  = c("simple"),
  num.simu   = 21:100,
  stringsAsFactors = FALSE
) 
```

```{r, eval=FALSE}
if (!dir.exists("results3")) dir.create("results3")

for (i in rows_along(params.grid3)) {

  res.file <- paste0("results3/simu_", i, ".rds")
  if (file.exists(res.file)) next

  params <- params.grid3[i, ]
  par.causal <- params[["par.causal"]][[1]]

  # Split in training/test sets
  ind.train <- sort(sample(n, size = n.train))
  ind.test <- setdiff(1:n, ind.train)

### Simulations n°4 (dataset with only chromosome 6)

```{r}
params.grid4 <- expand.grid(
  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = c(0.5, 0.8), 
  par.model  = c("simple"),
  num.simu   = 1:100,
  stringsAsFactors = FALSE
) 
```

@@

### Simulations n°5 (varying training size)

```{r}
params.grid5 <- expand.grid(
  n.train    = 1:5 * 1000,
  par.causal = list(c(300, "all")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = c(0.5, 0.8), 
  par.model  = c("simple"),
  num.simu   = 1:100,
  stringsAsFactors = FALSE
) 
```

>>>>

```{r, message=FALSE}
library(tidyverse)
library(glue)
```

@@

    cmsa.logit <- big_spLogReg(X = G, y01.train = pheno.all[ind.train], 
                               ind.train = ind.train, 
                               covar.train = covar.all[ind.train, , drop = FALSE],
                               ncores = nb_cores())

    preds <- rowMeans(
      predict(cmsa.logit, X = G, ind.row = ind.test, 
                     covar.row = covar.all[ind.test, , drop = FALSE])
    )
  })[3]

  tibble(
    method   = method, 
    eval     = list(cbind(preds, pheno.all[ind.test])),
    timing   = timing,
    nb.preds = sum(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0)
  )
}
```

@@

n <- nrow(G)
ind.HLA <- snp_indLRLDR(CHR, POS, subset(LD.wiki34, ID == "hild12"))
```

@@

### Scenario n°1 (with T-Trees)

```{r}
params.grid1 <- expand.grid(
  n.train    = 6000,
  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = 0.8, 
  par.model  = c("simple", "fancy"),
  num.simu   = 1:5,
  stringsAsFactors = FALSE
) 
```

```{r, eval=TRUE}
if (!dir.exists("results1")) dir.create("results1")

@@

@@

### Scenario n°1 (without T-Trees)

```{r}
params.grid2 <- expand.grid(
  n.train = 6000,
  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = c(0.5, 0.8), 
  par.model  = c("simple", "fancy"),
  num.simu   = 1:100,
  stringsAsFactors = FALSE
) 
```

```{r, eval=TRUE}
if (!dir.exists("results2")) dir.create("results2")

@@

### Scenario n°2 (dataset with only chromosome 6)

```{r}
params.grid4 <- expand.grid(
  n.train    = 6000,
  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = c(0.5, 0.8), 
  par.model  = "simple",
  num.simu   = 1:100,
  stringsAsFactors = FALSE
) 
```

@@

### Scenario n°3 (varying training size)

```{r}
params.grid5 <- expand.grid(
  n.train    = 1:5 * 1000,
  par.causal = list(c(300, "all")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = c(0.5, 0.8), 
  par.model  = "simple",
  num.simu   = 1:100,
  stringsAsFactors = FALSE
) 
```

### 2018-02-12 4-results-simus.Rmd
save

```{r}
format_results <- function(results) {
  results %>%
    as_tibble() %>%
    mutate(
      par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
                          levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
      AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
    )
}
```

@@

  results %>%
    myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
    geom_hline(yintercept = 0.5, linetype = 2) +
    geom_bar(stat = "identity", alpha = 0.3, position = position_dodge()) +
    geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
                  position = position_dodge(width = 0.9), color = "black", width = 0.2) + 
    scale_y_continuous(limits = c(0.5, NA), minor_breaks = 0:20 / 20,
                       oob = scales::rescale_none) +
    labs(x = "Causal SNPs (number and location)", y = "Mean of 100 AUCs",
         fill = "Method", color = "Method") +
    scale_fill_manual(values = methods.color) +
    scale_color_manual(values = methods.color)
}

@@

  nsimu <- results.summary$N[[1]]
  stopifnot(all(results.summary$N == nsimu))

@@

```{r, out.width="90%", fig.width = 14}
compare_logit(bind_rows(results1, results2), with_method = "logit-triple")

@@

```{r}
results2.summary %>%
  filter(grepl("logit", method)) %>%
  barplot_causal_one(h2 = 0.5) +
  facet_grid(par.dist ~ par.model) +
  labs(y = "Mean of 20 AUCs")

@@

```{r}
results2.summary %>%
  filter(par.model == "fancy", method %in% c("logit-simple", "PRS-max")) %>%
  barplot_causal_all() +
  labs(y = "Mean of 20 AUCs")

@@

```{r}
results2.summary %>%
  filter(par.model == "fancy", grepl("PRS", method)) %>%
  barplot_causal_all() +
  labs(y = "Mean of 20 AUCs")

@@

### Correlation between predictive measures

```{r}
results3 <- list.files(paste0("results", 2:3), full.names = TRUE) %>%
  map_dfr(~readRDS(.x)) %>%
  filter(method != "logit-triple", par.model != "fancy") %>%
  format_results()

results3_corr <- results3 %>%
  filter(method %in% c("logit-simple", "PRS-max")) %>%
  mutate(
    percCases10 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[1:110], 2])),
    percCases20 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[1:220], 2])),
    Parameters  = interaction(paste("h2 =", par.h2), par.dist, method, sep = " | ")
  )
```

```{r, fig.asp=0.5, out.width="90%", fig.width=15}
p1 <- results3_corr %>%
  myggplot(aes(AUC, percCases10, color = Parameters)) + 
  geom_point(size = 0.6, alpha = 0.5) +
  geom_smooth(aes(linetype = Parameters), method = "lm", se = FALSE, size = 2) +
  theme(legend.key.width = unit(3.5, "line"), 
        legend.direction = "horizontal") + 
  labs(y = "Proportion of cases among top 10% scores")
p2 <- results3_corr %>%
  myggplot(aes(AUC, percCases20, color = Parameters)) + 
  geom_point(size = 0.6, alpha = 0.5) +
  geom_smooth(aes(linetype = Parameters), method = "lm", se = FALSE, size = 2) + 
  labs(y = "Proportion of cases among top 20% scores") +
  theme(legend.position = "none")

cowplot::plot_grid(p1 + theme(legend.position = "none"), p2 , 
                   ncol = 2, align = "hv", scale = 0.95) %>%
  cowplot::plot_grid(cowplot::get_legend(p1), ., rel_heights = c(0.1, 1), ncol = 1)

ggsave("figures/supp-AUC-corr.pdf", scale = 1/90, width = 1558, height = 737)

with(results3_corr, cor(AUC, percCases10))
with(results3_corr, cor(AUC, percCases20))
```

@@

```{r, fig.asp = 1, out.width = "60%", fig.width=8}
results3.summary %>%
  filter(method %in% c("logit-simple", "PRS-max")) %>%
  barplot_causal_one(h2 = 0.8)

>>>>

```{r}
pAUC <- function(pred, target, p = 0.1) {
  val.min <- min(target)
  q <- quantile(pred[target == val.min], probs = 1 - p)
  ind <- (target != val.min) | (pred > q)
  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p, 
           error = function(e) p^2 / 2)  # all same prediction
}

format_results <- function(results) {
  results %>%
    as_tibble() %>%
    mutate(
      par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
                          levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
      AUC = map_dbl(eval, ~ bigstatsr::AUC(.x[, 1], .x[, 2])),
      pAUC = map_dbl(eval, ~ pAUC(.x[, 1], .x[, 2]))
    )
}
```

@@

  results %>%
    myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
    geom_hline(yintercept = 0.5, linetype = 2) +
    geom_col(position = position_dodge(), alpha = 0.3) +
    geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
                  position = position_dodge(width = 0.9), color = "black", width = 0.2) + 
    scale_y_continuous(limits = c(0.5, NA), minor_breaks = 0:20 / 20,
                       oob = scales::rescale_none) +
    labs(x = "Causal SNPs (number and location)", y = "Mean of 100 AUCs",
         fill = "Method", color = "Method") +
    scale_fill_manual(values = methods.color) +
    scale_color_manual(values = methods.color)
}

@@

  nsimu <- results.summary$N[[1]]
  # stopifnot(all(results.summary$N == nsimu))

@@

```{r, out.width="90%", fig.width = 14}
compare_logit(results2, with_method = "logit-triple")

@@

```{r}
results2.summary %>%
  filter(grepl("logit", method)) %>%
  barplot_causal_one(h2 = 0.5) +
  facet_grid(par.model ~ par.dist)

@@

```{r}
results2.summary %>%
  filter(par.model == "fancy", method %in% c("logit-simple", "PRS-max")) %>%
  barplot_causal_all()

@@

```{r}
results2.summary %>%
  filter(par.model == "fancy", grepl("PRS", method)) %>%
  barplot_causal_all()

@@

### Correlation between predictive measures [MOVE TOP]

```{r}
results3 <- results2 %>%
  filter(par.model == "simple", method != "logit-triple")
```

```{r, fig.asp=0.5, out.width="95%", fig.width=16}
results3 %>%
  filter(method %in% c("logit-simple", "PRS-max")) %>%
  mutate(Parameters  = interaction(par.h2, par.dist, method, sep = " | ")) %>%
  myggplot(aes(AUC, pAUC, color = Parameters)) + 
  geom_point(size = 0.6, alpha = 0.5) +
  geom_smooth(aes(linetype = Parameters), method = "loess", se = FALSE, size = 2) +
  theme(legend.key.width = unit(3.5, "line"), 
        legend.direction = "horizontal") + 
  labs(y = "Partial AUC (specificity: [0.9, 1])") +
  theme(legend.position = "top")

ggsave("figures/supp-AUC-corr.pdf", scale = 1/100, width = 1525, height = 983)
```

```{r}
cor(results3$AUC, results3$pAUC, method = "spearman")
```

@@

```{r, fig.asp = 1, out.width = "60%", fig.width=8}
results3.summary %>%
  filter(method %in% c("logit-simple", "PRS-max"), par.model == "simple") %>%
  barplot_causal_one(h2 = 0.8)

### 2018-02-12 paper/draft.tex
save

\newpage
\begin{table}[htbp]
\caption{Summary of all simulations. When there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.}
\vspace*{0.5em}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & Number of & \multirow{2}{*}{Methods} \\ 
scenario & & training set & (number and location) & of effects & & model & simulations & \\
\hline
\hline
\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{4}{*}{0.8} & \multirow{2.5}{*}{simple} & \multirow{4}{*}{5} & PRS \\
& & & 30 in all & & & & & logit-simple \\
& & & 300 in all & \multirow{1.5}{*}{Laplace} & & \multirow{1.5}{*}{fancy} & & logit-triple \\
& & & 3000 in all & & & & & T-Trees \\
\hline
\multirow{3}{*}{2} & \multirow{3}{*}{-} & \multirow{3}{*}{-} & \multirow{3}{*}{-} & \multirow{3}{*}{-} & \multirow{2}{*}{0.5} & \multirow{3}{*}{-} & \multirow{3}{*}{20} & PRS \\ 
& & & & & \multirow{2.2}{*}{0.8} & & & logit-simple \\
& & & & & & & & logit-triple \\
\hline
\multirow{2}{*}{3} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{simple} & \multirow{2}{*}{80} & PRS \\ 
& & & & & & & & logit-simple \\
\hline
\multirow{2}{*}{4} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{100} & \multirow{2}{*}{-} \\ 
& & & & & & & & \\
\hline
\multirow{5}{*}{5} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{100} & \multirow{5}{*}{-} \\ 
& & 2000 & & & & & & \\
& & 3000 & & & & & & \\
& & 4000 & & & & & & \\
& & 5000 & & & & & & \\
\hline
\end{tabular}
\label{tab:simus}
\end{table}

>>>>

\newpage
\begin{table}[htbp]
\caption{Summary of all simulations. When there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.}
\vspace*{0.5em}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
scenario & & training set & (number and location) & of effects & & model & \\
\hline
\hline
\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{simple} & PRS \\
& & & 30 in all & & & & logit-simple \\
& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{fancy} & logit-triple \\
& & & 3000 in all & & & & (T-Trees) \\
\hline
\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{simple} & PRS \\ 
& & & & & & & logit-simple\\
\hline
\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
& & 2000 & & & & & \\
& & 3000 & & & & & \\
& & 4000 & & & & & \\
& & 5000 & & & & & \\
\hline
\end{tabular}
\label{tab:simus}
\end{table}
